The Name of the Title Is Hope

### 2024-12-30 paper.tex
Push draft Background and Design

\section{Introduction}

\subsection{Goals of a Scan Implementation}
\begin{itemize}
  \item \textbf{Minimal global memory access:} the scan should minimize global memory access to the theoretical minimum of $2n$.
  \item \textbf{Fully saturates global memory bandwidth:} the scan should be capable of fully saturating the global memory bandwidth.
  \item \textbf{Portability:} the scan should be portable across hardware vendors and architectures. 
\end{itemize}

@@

\subsection{The Scan Primitive}
The study of \emph{scan} (\emph{parallel prefix}) networks traces back to the design of carry-lookahead adder circuits and beyond\cite{10.1145/322217.322232, 5219801}. A scan is typically defined on a monoid \( M \), characterized by a binary reduction operator \( \oplus \) and an identity element \( \varepsilon \). The binary operator \( \oplus \) satisfies the closure property \( \forall a, b \in M, \ (a \oplus b) \in M \) and has an identity element \( \exists \varepsilon \in M, \ \forall a \in M, \ \varepsilon \oplus a = a \). Although \( \oplus \) must be associative, it is not necessarily commutative, as demonstrated in structures like the stack monoid\cite{}. In a scan, the result at the \( n \)-th element is the reduction of the preceding subset of elements in the sequence. If the subset includes the \( n \)-th element, it is called \textit{inclusive}; if it excludes the \( n \)-th element, it is called \textit{exclusive}. The most common scan type is the prefix sum, where \( \oplus \) is addition. For example:
\[
  x = [x_1, x_2, x_3, \dots, x_n] \ \ \ \ \ \ \ y = [1, 1, 1, 1, 1] 
\]
\[
  \text{InclusiveScan}(x, \oplus) = [x_1, x_1 \oplus x_2, x_1 \oplus x_2 \oplus x_3, \dots, x_1 \oplus x_2 \oplus \cdots \oplus x_n]
\]
\[
  \text{InclusiveScan}(y, +) = [1, 2, 3, 4, 5]
\]

Due to its significance in circuits and as a fundamental algorithmic primitive, scan enjoys a rich literature in both electrical engineering and computer science~\cite{}. As comprehensive taxonomies of scan networks are well-documented, our focus is on exploring the progression of inter-workgroup (global) scan strategies and their fitness relative to our ideal architecture.

@@

\subsubsection{Scan-then-Propagate}
Introduced by Sengupta et al.\cite{10.5555/1280094.1280110}, the \emph{Scan-then-Propagate} \cite{GPUGems3, Sengupta2008} strategy is an intuitive extension of the intra-workgroup scan, comprising three phases: 
\begin{enumerate} 
  \item \textbf{Intra-Workgroup (Local) Scan:} Each workgroup performs an inclusive scan on its assigned work-tile partition of the input data. The results, including both the local scan output and the reduction of each workgroup, are written back to global memory.
  \item \textbf{Spine Scan:} The reductions from all workgroups, collectively referred to as \emph{the spine}, are gathered and scanned to compute the global prefix of the reductions, resolving the dependencies between workgroups; we call the result of this phase the \emph{root}. 
  \item \textbf{Propagation:} The root is propagated back to each workgroup and combined with the local scan results to compute the final scan. 
\end{enumerate}
Without an inter-workgroup barrier primitive, each phase is implemented as a separate kernel launch to ensure a consistent view of the reductions. During the scan operation, each element is read from and written to global memory once during the local scan and again during the propagation phase, contributing a total of $4n$ global memory operations. Although data movement is dominated by the first and third phase, Sengupta et al.'s implementation handles arbitrarily large inputs using recursion. For an input of size $n$ and a work tile of size $t$, this results in a recursive depth of $\lceil \log_t n \rceil$. At each recursive step $k$, $n/t^k$ tiles are processed, leading to a total of $\sum_{k=1}^{\lceil \log_t n \rceil} n/t^k \approx (n - 1)/(t - 1)$ work tiles over all steps. Since each tile moves $4t$ data ($2t$ local scan, $2t$ propagation), the total data movement is $O\left(\frac{4t(n - 1)}{t - 1}\right) = O(4n)$.

\subsubsection{Reduce-then-Scan}
As the design of GPUs settled around the high-throughput but high-latency paradigm, reducing global memory access, rather than compute, became the prevailing goal for algorithms like scan, which are computationally light enough for their compute to be masked by global memory access. This led to a refinement of the \emph{Scan-Then-Propagate} strategy called \emph{Reduce-Then-Scan}\cite{Merrill Grimshaw, Ha and Han, Dotsenko, Breitbart}: by writing only the work tile reduction in the first phase and performing a redundant scan in the third phase, $n$ global data movement can be eliminated. The \emph{Reduce-Then-Scan} strategy operates in three phases:
\begin{enumerate}
  \item \textbf{Work Tile Reduction:} Each workgroup performs a pure reduction on its work tile partition, posting the result into global memory. 
  \item \textbf{Spine Scan:} The reductions from all workgroups are gathered and scanned to produce the root.
  \item \textbf{Intra-Workgroup Scan and Propagation:} Each workgroup performs the local scan on the work tile, incorporating the root as it writes to global memory.
\end{enumerate}
\emph{Reduce-then-Scan} implementations also introduced improvements to the spine phase of the scan, which coalesced around a workgroup-level \emph{raking} technique first applied to inter-workgroup scan by Merrill and Grimshaw~\cite{}. Instead of handling arbitrary inputs recursively, the total workgroup occupancy $o$ is determined, and the input is partitioned into non-overlapping blocks processed serially by each workgroup. This approach limits the recursive depth of the global scan to two and reduces the global memory movement for the spine scan from $O(n/t)$ (incurred by recursion) to $O(3o)$. As a result, kernel launches are limited to three and total global memory movement is reduced to $O(3n + 3o) = O(3n)$.

\subsubsection{Chained-Scan}
Continuing the trend of global data movement optimization, the most recent work on global scan has been the \emph{Chained-Scan} architecture. First introduced by Yan et al. in \emph{StreamScan}, the key innovation of GPU chained scanning lies in its hybridization of parallel and serial strategies: achieving parallelism at the intra-workgroup level while minimizing global data movement through serial scan operations at the inter-workgroup level.

Instead of using kernel launches as inter-workgroup synchronization points, \emph{Chained-Scans} launch a single kernel that assigns work tiles in a serial order to workgroups. This shifts the problem from explicit synchronization to implicit synchronization via the order in which workgroups access global data. Each workgroup processes its assigned tile independently, waiting only for the result of the preceding reductions before continuing. This structure ensures that global memory traffic is minimized, as each element is read and written exactly once, achieving the theoretical minimum of $2n$.

While \emph{StreamScan} successfully reduces per-element device memory accesses to two, it does not consistently achieve the expected throughput of $\text{memoryBandwidth}/2$ (saturate bandwidth?). This limitation arises because, in \emph{StreamScan}, a workgroup cannot post its reduction to global memory until it has ingested the reduction of the preceding workgroup. This sequential dependency creates a bottleneck, constraining the algorithm’s performance to the rate at which reductions can propagate through device memory—a process limited by the relatively high latency of GPU memory.

Building on \emph{StreamScan}, Merrill and Garland introduced the \emph{Decoupled Lookback} technique to address the bottleneck created by these sequential dependencies. Unlike \emph{StreamScan}, where reductions and dependency resolution occur sequentially, \emph{Decoupled Lookback} separates these processes.In \emph{Decoupled Lookback}, each workgroup posts its reduction to device memory and then enters the “lookback” phase. During this phase, the workgroup performs a serial traversal backward through preceding workgroup reductions stored in device memory. To manage inter-workgroup synchronization, bit-packed status flags are used to indicate the state of each workgroup’s reduction.

TODO Cleanup, add O(n) on global data movement during the lookback phase etc.

\subsection{Ideal Scan Architecture}
RTS
More portable
slower
handling arbitrary inputs can be awkward

Chained Scan
Faster
2n data movement + serial ordering faclitates in-situ behavior
naturally extends to arbitrary data
not portable, relies on something called forward progress guarantees

Chained has lots of advantages, it would be great if it was portable.

\subsection{Why does \emph{Chained-Scan} Rely on Forward Progress Guarantees?}
TODO rewrite
The \emph{Chained-Scan} architecture operates in the middle ground between the absence of inter-workgroup synchronization and formal inter-workgroup synchronization support. Rather than relying on explicit synchronization primitives, it leverages a combination of atomic operations and a property of the hardware's workgroup scheduling model known as a \emph{forward progress guarantee}. Atomics enable coherent communication between workgroups, while the \emph{forward progress guarantee} (FPG) ensures that all active workgroups will eventually make progress towards termination, precluding scenarios where a workgroup is scheduled in a manner that starves another workgroup of execution time.

In \emph{Chained-Scan}, workgroups operate within a producer-consumer framework: each workgroup produces the reduction of its assigned work tile, and every workgroup (except the first) must consume the reductions produced by all preceding workgroups to complete the scan. To guarantee that no workgroup begins scanning a tile for which the preceding reduction may be unavailable, \emph{Chained-Scan} assigns work tiles using atomic increment operations rather than assignment based on workgroup index. Once a workgroup acquires its tile, it computes the tile’s reduction, atomically posts the result into global memory, and then waits for preceding reductions to appear in global memory. It is this waiting phase that necessitates FPG. Because workgroups do not execute in physical lockstep (obvious?), some may begin waiting before others have finished their calculations. Without FPG, there is a risk that a workgroup which finishes early could remain indefinitely scheduled in a waiting state, blocking progress by the workgroup responsible for producing the needed reduction and ultimately causing a deadlock.

As an illustrative example, consider a scan operation which requires exactly two workgroups, running on a GPU with a single multiprocessor. While this multiprocessor can host both workgroups, it contains only a single scheduling unit, which lacks FPG. In this scenario, both workgroups launch and acquire their respective work tiles, but the workgroup responsible for \emph{Work Tile 1} finishes first and begins waiting on \emph{Work Tile 0}. Because the scheduler lacks FPG, it continues scheduling \emph{Work Tile 1}, but because there is only a single scheduling unit, \emph{Work Tile 0} is never scheduled and never completes its reduction. As a result, \emph{Work Tile 1} spins indefinitely until \emph{timeout detection and recovery} (TDR) is triggered, crashing the host program in the process.

\subsection{Why is reliance on Forward Progress Guarantees a Portability Problem?}
Although NVIDIA formalized FPG down to the subgroup level beginning with the Volta architecture---and FPG was likely already present at the workgroup level on at least NVIDIA's Fermi and AMD's TeraScale2 architectures~\cite{}---it remains absent on some contemporary hardware, most notably certain ARM-based chips, including Apple's M-series GPUs~\cite{10.1145/3485508}. Our testing shows that, at best, attempting to run \emph{Chained-Scan} without FPG results in mega-bad\footnote{Need to gather data on the distribution of M1 Pro time on Decoupled Lookback instead of the average time. It's also unclear why the test did not trigger Metal's timeout device recovery}, and at worst, risks the afformentioned TDR and subsequent program crash. This risk is further exacerbated by the fact that, no major graphics API currently provides developers an FPG hardware capability query\footnote{To the best of our knowledge D3D12, Vulkan, and Metal do not have a query for FPG}. As a result, when implementing scan operations in a shading environment—where cross-vendor, cross-architecture portability is essential—uncertainty surrounding FPG capability and its catastrophic failure mode compels developers to fall back to the older, slower \emph{Reduce-then-Scan} approach.

\subsection{Earlier Attempts at Portability}
Scalar fallback here.

\section{Design}
\subsection{Goals (These subsection headers can be removed later if necessary)}
This work is guided by two main goals. First, portability: we want to develop a scan implementation that retains the benefits of the \emph{Chained-Scan} architecture but is also suitable for a diverse range of hardware vendors and achitectures, including those without FPG. Second, performance: we aim for near speed-of-light execution to the greatest extent possible allowed by the underlying hardware and programming model. To ground these objectives, we select the WebGPU shading language (WGSL) and the WebGPU standard as our target environment. WGSL is a meta-level shading language which is translated and compiled as necessary for backend graphics APIs---D3D12, Vulkan, and Metal---and as such, WGSL implementations are bound by the minimum capability across all backends. Because WGSL must operate within this limited capability space, it inherently embodies the challenge of portability.

@@

\subsection{Achieving Goals}

@@

\subsection{Intra-Workgroup Implementation}
Our intra-workgroup implementation is most similar to the \emph{CUDPP} scan kernel \cite{}, with adaptations to enhance portability. It consists of three phases:
\begin{enumerate}
  \item
\end{enumerate}

>>>>

\section{Introduction}
(Tell readers we are targetting WebGPU)
(Speed of light definition here)
\subsection{Goals of a Scan Implementation}
\begin{itemize}
  \item \textbf{Minimal global memory access:} the scan should minimize global memory access to the theoretical minimum of $2n$.
  \item \textbf{Fully saturates global memory bandwidth:} the scan should be capable of fully saturating the global memory bandwidth.
  \item \textbf{Portability:} the scan should be portable across hardware vendors and architectures. 
\end{itemize}

@@

\subsection{The Scan Primitive}
The study of \emph{scan} (\emph{parallel prefix}) networks traces back to the design of carry-lookahead adder circuits and beyond~\cite{10.1145/322217.322232, 5219801}. A scan is typically defined on a monoid \( M \), characterized by a binary reduction operator \( \oplus \) and an identity element \( \varepsilon \). The binary operator \( \oplus \) satisfies the closure property \( \forall a, b \in M, \ (a \oplus b) \in M \) and has an identity element \( \exists \varepsilon \in M, \ \forall a \in M, \ \varepsilon \oplus a = a \). Although \( \oplus \) must be associative, it is not necessarily commutative, as demonstrated in structures like the stack monoid~\cite{}. In a scan, the result at the \( n \)-th element is the reduction of the preceding subset of elements in the sequence. If the subset includes the \( n \)-th element, it is called \textit{inclusive}; if it excludes the \( n \)-th element, it is called \textit{exclusive}. The most common scan type is the prefix sum, where \( \oplus \) is addition. For example:
\[
  x = [x_1, x_2, x_3, \dots, x_n] \ \ \ \ \ \ \ y = [1, 1, 1, 1, 1] 
\]
\[
  \text{InclusiveScan}(x, \oplus) = [x_1, x_1 \oplus x_2, x_1 \oplus x_2 \oplus x_3, \dots, x_1 \oplus x_2 \oplus \cdots \oplus x_n]
\]
\[
  \text{InclusiveScan}(y, +) = [1, 2, 3, 4, 5]
\]

Due to its significance in circuits and as a fundamental algorithmic primitive, scan enjoys a rich literature in both electrical engineering and computer science. Harris provides a taxonomy relating depth, fanout, and wire tracks. Hinze provides an algebra of scans. Snir proved that depth and size are related by . Fich proved that . Merill and Garland provide a comprehensive review of GPU scan implementations in . In light of this work(reword this), our focus is on exploring the progression of inter-workgroup (global) scan strategies and their fitness relative to our ideal architecture.

Due to its significance in circuits and as a fundamental algorithmic primitive, scan has been extensively studied in both electrical engineering and computer science. Harris offers a taxonomy that relates depth, fanout, and wire tracks~\cite{}, while Hinze develops an algebraic framework for scans~\cite{}. Snir proved that \emph{depth} and \emph{size} are related by (fill in)~\cite{}, and Fich proved that among minimum depth scans Ladner-Fischer networks have optimal size. Finally, Merrill and Garland provide a comprehensive review of GPU scan implementations. In this work, we focus on the evolution of inter-workgroup (global) scan strategies and evaluate their suitability for our ideal architecture.

@@

\subsubsection{Scan-then-Propagate}
Introduced by Sengupta et al.~\cite{10.5555/1280094.1280110}, the \emph{Scan-then-Propagate} ~\cite{GPUGems3, Sengupta2008} strategy is an intuitive extension of the intra-workgroup scan, comprising three phases: 
\begin{enumerate} 
  \item \textbf{Intra-Workgroup (Local) Scan:} Each workgroup performs an inclusive scan on its assigned work-tile partition of the input data. The results, including both the local scan output and the reduction of each workgroup, are written back to global memory.
  \item \textbf{Spine Scan:} The reductions from all workgroups, collectively referred to as \emph{the spine}, are gathered and scanned to compute the global prefix of the reductions, resolving the dependencies between workgroups; we call the result of this phase the \emph{root}. 
  \item \textbf{Propagation:} The root is propagated back to each workgroup and combined with the local scan results to compute the final scan. 
\end{enumerate}
Without an inter-workgroup barrier primitive, each phase is implemented as a separate kernel launch to ensure a consistent view of the reductions. During the scan operation, each element is read from and written to global memory once during the local scan and again during the propagation phase, contributing a total of $4n$ global memory operations. Although data movement is dominated by the first and third phase, Sengupta et al.'s implementation handles arbitrarily large inputs using recursion. For an input of size $n$ and a work tile of size $t$, this results in a recursive depth of $\lceil \log_t n \rceil$. At each recursive step $k$, $n/t^k$ tiles are processed, leading to a total of $\sum_{k=1}^{\lceil \log_t n \rceil} n/t^k \approx (n - 1)/(t - 1)$ work tiles over all steps. Since each tile moves $4t$ data ($2t$ local scan, $2t$ propagation), the total data movement is $O\left(\frac{4t(n - 1)}{t - 1}\right) = O(4n)$.

\subsubsection{Reduce-then-Scan}
Due to the extreme latency of global memory access relative to compute, computationally light algorithms like \emph{scan} can benefit from redundant calculation if it reduces memory access. This insight led to a refinement of the \emph{Scan-Then-Propagate} strategy called \emph{Reduce-Then-Scan}~\cite{Merrill Grimshaw, Ha and Han, Dotsenko, Breitbart}: by writing only the work tile reduction in the first phase and performing a redundant scan in the third phase, $n$ global data movement can be eliminated. The \emph{Reduce-Then-Scan} strategy operates in three phases:
\begin{enumerate}
  \item \textbf{Work Tile Reduction:} Each workgroup performs a pure reduction on its work tile partition, posting the result into global memory. 
  \item \textbf{Spine Scan:} The reductions from all workgroups are gathered and scanned to produce the root.
  \item \textbf{Intra-Workgroup Scan and Propagation:} Each workgroup performs the local scan on the work tile, incorporating the root as it writes to global memory.
\end{enumerate}
Merrill and Grimshaw~\cite{} introduced a workgroup-level \emph{raking} technique that improved the spine phase of \emph{Reduce-Then-Scan}. Instead of handling arbitrary inputs recursively, this approach determines the total workgroup occupancy $o$ and partitions the input into non-overlapping blocks, processed serially by each workgroup. By limiting the recursive depth of the global scan to two, the global memory movement for the spine scan is reduced from $O(n/t)$ to $O(3o)$. Consequently, kernel launches are limited to three, and total global memory movement is reduced to $O(3n + 3o) = O(3n)$. This approach was further refined by Ha and Han~\cite{}, who modified the upsweep behavior of the first workgroup from reduction to scan and eliminated the redundant upsweep of the last workgroup, thereby saving the equivalent of one workgroup's computation.

\subsubsection{Chained-Scan}
Continuing the trend of global data movement optimization, the most recent work on global scan has been the \emph{Chained-Scan} architecture. First introduced by Yan et al. in \emph{StreamScan}, the key innovation of GPU chained scanning lies in its hybridization of parallel and serial strategies: achieving parallelism at the intra-workgroup level while minimizing global data movement through serial scan operations at the inter-workgroup level. Instead of using kernel launches as inter-workgroup synchronization points, \emph{Chained-Scans} launch a single kernel that assigns work tiles in a serial order to workgroups. This shifts the problem from explicit synchronization to implicit synchronization via the order in which workgroups access global data. Each workgroup processes its assigned tile independently, waiting only for the result of the preceding reductions before continuing. This structure ensures that global memory traffic is minimized, as each element is read and written exactly once, achieving the theoretical minimum of $2n$.

While \emph{StreamScan} successfully reduces per-element device memory accesses to two, it does not consistently saturate global memory bandwidth, preventing it from achieving speed-of-light performance. This limitation arises because a workgroup in \emph{StreamScan} cannot post its reduction to global memory until it has received the reduction of the preceding workgroup. Although this reduces global memory access during the propagation phase to $O(n/t)$, the sequential dependency creates a bottleneck, constraining the algorithm’s performance to the rate at which reductions propagate through device memory. Despite this limitation, the total data movement remains optimal at $O(2n + n/t) = O(2n)$.

Building on \emph{StreamScan}, Merrill and Garland introduced the \emph{Decoupled Lookback} technique to address the bottleneck created by sequential dependencies. Unlike \emph{StreamScan}, where reductions and dependency resolution occur sequentially, \emph{Decoupled Lookback} separates these processes. In this approach, each workgroup posts its local reduction to global memory and then performs a \emph{lookback}—a serial traversal backward along the spine to resolve dependencies. The work performed during this phase is constant-bounded to the workgroup occupancy $o$, as redundant operations are avoided by updating the workgroup's reduction from local to inclusive once the lookback is completed. Although this increases global memory access during the propagation phase to $O(o(n/t))$, the serial ordering of work tile processing ensures that memory accesses to the spine are roughly contiguous within the workgroup occupancy block, making them highly likely to be cache-resident. As a result, the additional global memory traffic along the spine introduces a negligible latency overhead. With a total global memory movement of $O(2n+o(n/t))= O(2n)$, \emph{Decoupled Lookback} achieves full utilization of device memory bandwidth and delivers speed-of-light performance.

\subsection{Ideal Scan Architecture}

In addition to its speed, the \emph{Chain-Scan} architecture has other, more practical properties, that improve over its predecessors. As it is a single kernel, it minimizes launch overheads and simplifies maintenance. On previous architectures, handling of arbitrary inputs can be awkward: the recursive approach, for example, entails multiple kernel launches and a heterogenous system to coordinate them. Similarly, a raking approach can be cumbersome, as determining the workgroup occupancy for every target device is often impractical. Furthermore, on GPUs where the global memory data cache is indexed by physical addresses—and therefore requires lookups in \emph{translation lookaside buffers} (TLBs)—raking the workgroups can lead to performance degradation due to cache thrashing. This issue occurs when task sizes are large enough that each workgroup accesses its own unique page entry, and the device lacks sufficient TLB coverage to support the memory access patterns of all workgroups. As a consequence of its $2n$ global data movement and inherently serial ordering, \emph{Chained-Scan} is also suitable for \emph{in-situ} compaction and allocation. In previous architectures, such tasks often required additional kernel launches or complex bookkeeping to handle dependencies between workgroups. By contrast, \emph{Chained-Scan} efficiently propagates dependencies within a single kernel, minimizing overhead and enabling seamless integration of compaction and allocation operations.

Thus, while \emph{Chained-Scan} with \emph{Decoupled Lookback} appears to be the ideal scan architecture, it falls short in meeting the criterion of \emph{portability}. As we discuss in more detail in the following section, this limitation arises because \emph{Chained-Scan} depends on a \emph{forward-progress guarantee}, a feature provided by NVIDIA hardware but not universally available across all vendors and architectures. Without this guarantee, \emph{Chained-Scan} is highly prone to catastrophic failure, making it unsuitable for the inherently cross-vendor, cross-architecture environment of shading languages.

\subsection{Why does \emph{Chained-Scan} Rely on Forward-Progress Guarantees?}
A \emph{forward-progress guarantee} (FPG) is a promise provided by the subgroup scheduler that all occupant subgroups eventually make progress toward termination, preventing scenarios where a subgroup is starved of execution time. Sorensen et al.~\cite{} and Sorensen et al.~\cite{} provide the most comprehensive analysis of GPU schedulers, including the implications of running FPG dependent algorithms without it.

In \emph{Chained-Scan}, workgroups operate in a producer-consumer framework: each workgroup produces the reduction of its assigned work tile, while subsequent workgroups consume the reductions of all preceding workgroups to complete the scan. To enforce this serial ordering, work tiles are dynamically assigned using atomic increment operations, rather than static assignment based on virtualized workgroup indices. While this dynamic assignment guarantees that predecessor work tiles are either scheduled or completed, without FPG there is no guarantee that these reductions will eventually become available. A scheduler lacking FPG may fail to allocate sufficient execution time to certain workgroups, leaving their reductions incomplete indefinitely. In such cases, consuming workgroups stall during their lookback phase, waiting for reductions that are never posted to global memory.

For instance, consider a scan operation requiring two workgroups running on a GPU with a single multiprocessor. Although the multiprocessor can host both workgroups, it has only one scheduling unit, which lacks FPG. In this setup, both workgroups launch and acquire their respective work tiles. The workgroup responsible for \emph{Work Tile 1} finishes first and begins waiting for the reduction of \emph{Work Tile 0}. However, because the scheduler lacks FPG, it may continue scheduling \emph{Work Tile 1}, leaving \emph{Work Tile 0} starved and unable to complete its reduction. As a result, \emph{Work Tile 1} spins indefinitely, eventually triggering \emph{timeout detection and recovery} (TDR) and crashing the host program.

\subsection{Why is reliance on Forward Progress Guarantees a Portability Problem?}
Although NVIDIA formalized FPG down to the intra-subgroup level beginning with the Volta architecture---and FPG was likely already present at the workgroup level on at least NVIDIA's Fermi and AMD's TeraScale2 architectures~\cite{}---it remains absent on some contemporary hardware, most notably certain ARM-based chips, including Apple's M-series GPUs~\cite{10.1145/3485508}. Our testing shows that, at best, attempting to run \emph{Chained-Scan} without FPG results in mega-bad\footnote{Need to gather data on the distribution of M1 Pro time on Decoupled Lookback instead of the average time. It's also unclear why the test did not trigger Metal's timeout device recovery}, and at worst, risks the afformentioned TDR and subsequent program crash. This extreme slowdown highlights the inherent fragility of \emph{Chained-Scan} without FPG.

Exacerbating this issue is the fact that no major graphics API currently provides developers with a mechanism to query FPG support\footnote{As of writing this, 12/29/2024, we are unaware of any major graphics API---Vulkan, D3D12, or Metal---that provide developers with an FPG capability query}. This lack of transparency forces developers to design algorithms without clear knowledge of whether critical guarantees are available on the target hardware. In environments like shading languages, where cross-platform portability is required, this uncertainty compels developers to revert to older, less efficient strategies such as \emph{Reduce-Then-Scan}. However, as discussed earlier, \emph{Reduce-Then-Scan} carries its own set of drawbacks, including increased global memory traffic ($O(3n)$) and additional kernel launches, which result in reduced throughput and higher latency. While these tradeoffs were once deemed acceptable, they stand in stark contrast to the efficiency and simplicity of \emph{Chained-Scan} when implemented on hardware with reliable FPG. Yet, in the absence of guaranteed forward progress, developers have little choice but to sacrifice performance for broader compatibility.

\subsection{Earlier Attempts at Portability (Cuttable?)}
Levien and Naur~\cite{}, made the first attempt at solving forward progress in \emph{DecoupledLookback}, introducing \emph{ScallarFallback}. In \emph{ScallarFallback}, each time a workgroup waits on a preceding tile, it begins a \emph{fallback} operation, and begins consuming the blocking tile one element at a time. Although this entails performing redundant work when the blocking tile is simply late but not deadlocking, it guarantees the termination of the algorithm because a workgroup can also perform the reduction of a blocking tile itself, thus breaking the inter-workgroup dependency.

\emph{ScallarFallback} suffers from two primary issues that impact its efficiency. First, the fallback reduction is performed serially by a single thread, progressing only when the workgroup engages in a waiting spin.This effectively takes $t$ spins to complete the fallback of the deadlocking tile. Moreover, because only a single thread is responsible for loading elements, the cache line is severely underutilized, further degrading performance. Second, once the fallback is complete, the workgroup does not post its result to device memory. Consequently, every workgroup deadlocked by that tile must redundantly execute the fallback, wasting computational resources. While \emph{ScallarFallback} was able to achieve speed of light on hardware supporting FPG and executed correctly on hardware without FPG, its performance on the latter was inferior to that of \emph{Reduce-Then-Scan} .

\section{Design}
\subsection{Goals}
This work is guided by two main goals. First, portability: we want to develop a scan implementation that retains the benefits of the \emph{Chained-Scan} architecture but is also suitable for a diverse range of hardware vendors and achitectures, including those without FPG. Second, performance: we aim for near speed-of-light execution to the greatest extent possible allowed by the underlying hardware and programming model. To ground these objectives, we select the WebGPU shading language (WGSL) and the WebGPU standard as our target environment. WGSL is a meta-level shading language which is translated and compiled as necessary for backend graphics APIs---D3D12, Vulkan, and Metal---and as such, WGSL implementations are bound by the minimum capability across all backends. Because WGSL must operate within this limited capability space, it inherently embodies the challenge of portability.

@@

@@

\subsection{Intra-Workgroup Implementation}
Our intra-workgroup implementation is most similar to the \emph{CUDPP} scan kernel ~\cite{}, with adaptations to enhance portability. It consists of three phases:
\begin{enumerate}
  \item
\end{enumerate}
